### 2017-05-07 2017mergingmerges/section/02-Processes.tex
Edits to S2

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure*}
\begin{alltt}
       merge
         = \(\lambda\) (sIn1: Stream Nat) (sIn2: Stream Nat) (sOut2: Stream Nat). 
           \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
           process
            { ins:    { sM1, sM2 }
            , outs:   { sM3 }
            , heap:   { x1 = 0, x2 = 0 }
            , label:  B0
            , instrs: { B0 = pull sIn1  x1   B1 []             , B1 = pull sIn2  x2   C0 []
                      , C0 = case (x1 < x2)  D0 []  E0 []      , D0 = push sOut2 x1   D1 []
                      , D1 = drop sIn1       D2 []             , D2 = pull sIn1  x1   C0 []
                      , E0 = push sOut2 x2   E1 []             , E1 = drop sIn2       E2 []
                      , E2 = pull sIn2 x2    C0 [] } }
\end{code}
% \begin{minipage}{0.29\textwidth}
\includegraphics[scale=1.0]{figures/state-merge.pdf}
% \includegraphics[scale=1.0,angle=-90,origin=c]{figures/state-merge.pdf}
% \end{minipage}

\caption{The merge combinator}
\label{fig:Process:Merge}
\end{figure*}

@@

We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams. For the @F0@ case shown in Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

@@

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transition from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.

@@

% The stream buffer variables we additionally introduced for coordination were freshly created during fusion.

% -----------------------------------------------------------------------------
\subsection{Breaking It Down}
We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. Importantly, we needed to \emph{break down} the definition of each of the operators into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, in the fused result program we have performed the job of a concurrent scheduler at compile time. The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.

The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion~\cite{lippmeier2013data}. However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss in \S\ref{s:FutureWork}.

Note that trying to use lazy evaluation for an example such as @uniquesUnion@ does not solve the buffering problem. Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists. As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirity before demanding any of the elements from the second list (@sUnion@). If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before remanding any of @sIn2@, causing a space leak. Lazy evaluation is \emph{pull only} meaning that evaluation is driven by the consumer. The space efficiency of our fused program relies critically on the fact that operators can also \emph{push} their result values directly to their consumers, and that the consumers are not at liberty to defer handling of these values.

@@

% -----------------------------------------------------------------------------
% BL: This is too much low level detail at this point, we need more intuition.

% The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.

% The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@. These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream. Thus the @group@ instructions have been duplicated, and are performed once when @merge@ initialises, and again every time @merge@ pulls from the first stream.

% Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions. As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop. In the fused labels on the right hand column, the other process still has @s1@ as `pending' or `have', so the element cannot yet be dropped. The actual @drop@ is only performed once both processes are `none', at label @F9@.

>>>>

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure*}
\begin{alltt}
       merge
         = \(\lambda\) (sIn1: Stream Nat) (sIn2: Stream Nat) (sOut2: Stream Nat). 
           \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
           process
            { ins:    { sM1, sM2 }
            , outs:   { sM3 }
            , heap:   { x1 = 0, x2 = 0 }
            , label:  B0
            , instrs: { B0 = pull sIn1  x1   B1 []             , B1 = pull sIn2  x2   C0 []
                      , C0 = case (x1 < x2)  D0 []  E0 []      , D0 = push sOut2 x1   D1 []
                      , D1 = drop sIn1       D2 []             , D2 = pull sIn1  x1   C0 []
                      , E0 = push sOut2 x2   E1 []             , E1 = drop sIn2       E2 []
                      , E2 = pull sIn2 x2    C0 [] } }
\end{code}
\includegraphics[scale=1.0]{figures/state-merge.pdf}
\caption{The merge combinator}
\label{fig:Process:Merge}
\end{figure*}

@@

We also associate each of the joint result states with an description of whether each source machine has already pulled a value from each of its input streams. For the @F0@ case shown in Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

@@

Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.

@@

% -----------------------------------------------------------------------------
\subsection{Breaking It Down}
We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. Importantly, we needed to \emph{break down} the definition of each operator into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, during fusion we have performed the job of a concurrent scheduler at compile time. The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching --- as there would be in a standard concurrent implementation using multiple threads.

The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative statements that implement each operator was also used in prior work on Flow Fusion~\cite{lippmeier2013data}. However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams.

Note that relying on lazy evaluation for an example such as @uniquesUnion@ does not solve the buffering problem. Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists. As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirity before demanding any of the elements from the second list (@sUnion@). If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before demanding any of @sIn2@, causing a space leak. Lazy evaluation is \emph{pull only} meaning that evaluation is driven by the consumer. The space efficiency of our fused program relies critically on the fact that operators can also \emph{push} their result values directly to their consumers, and that the consumers cannot defer the handling of these values.

@@
